---
title: "Estimating Properties of the Exponential Distribution from Random Samples"
author: "John Cartmill"
date: "February 2014"
output: 
        pdf_document:
                fig_caption: true

---


### Overview  
This paper will use **R** statistical package to apply the principles of the Central Limit Theorem to the exponential distribution to estimate the mean of the distribution.  
The exponential distribution with rate $lambda (\lambda)$ has density: 
$$ f(x) = \lambda e^{-\lambda x} for x \geq 0. $$  The mean value of the distribution is given by 1/\(\lambda\). The standard deviation, \(\sigma\), is also 1/\(\lambda\).  



```{r, echo=FALSE,warning=FALSE}
library(ggplot2)
library(xtable)
summaryfunction= function (x){
  if( is.numeric(x)!=TRUE) {stop("Supplied X is not numeric")}
  mysummary = data.frame(
            "Min." =as.numeric( min(x)),
            "1st Qu." = quantile(x)[2],
            "Median" = median(x),
            "Mean" = mean(x),
            "3rd Qu." = quantile(x)[4],
            "Max." = max(x),
            row.names=""
              
            )
  names(mysummary) = c("Min.","1st Qu.","Median","Mean","3rd Qu.","Max.")
  return( mysummary )
  }
```

```{r exp_dist, echo=FALSE, fig.height=3, fig.width=6,fig.cap="The exponential distribution for two values of lambda"}
k <- 1000
xvals <- seq(-1, 20, length = k)
myplot <- function(lambda) {
d <- data.frame(y = c(dexp(xvals,lambda ),dexp(xvals, lambda*2)),
x = xvals,dist = factor(rep(c(as.character(lambda), as.character(lambda * 2)), c(k,k))))
g <- ggplot(d, aes(x = x, y = y)) 
#g<- g + ggtitle("Exponential Distribution")
 g <- g + geom_line(size = 1, aes(colour = dist))+theme_classic(base_size=12) 
g<- g + scale_colour_discrete(name  ="lambda")
g <- g + theme(axis.title.x = element_text( colour="#000000", size=14),axis.text.x  = element_text(angle=90, vjust=0.5, size=16))
 g <- g + theme(axis.title.y = element_text( colour="#000000", size=14),axis.text.y  = element_text(angle=00, vjust=0.5, size=16)) + ylab("f(x)")
g<- g + annotate("text", x = 7.5, y = dexp(5,0.2)*1.1, label = "mean = 5",colour = "#8080FF")
g<- g +  annotate("segment", x = 5, xend = 5, y = -0.01, yend = dexp(5,0.2),colour = "#8080FF", size=1.0,linetype="twodash")
g<- g + annotate("text", x = 12.5, y = dexp(10,0.1)*1.2, label = "mean = 10",colour = "#FF8080")
g<- g +  annotate("segment", x = 10, xend = 10, y = -0.01, yend = dexp(10,0.1), colour = "#FF8080", size=1.0,linetype="twodash")

g
}
myplot(0.1)
```
  
In this paper all of the calculations were performed with  \(\lambda\) = 0.2.
For this choice of \(\lambda\) the mean is 5 and the variance(\(\sigma^2\)) is 25. To estimate the mean, 1000 sets of 40 samples were used.  

Here is the basic **R** code or the used to generate the sample sets:  
set.seed(7880)  
for (i in 1 : 1000) {  
        mns = c(mns, mean(rexp(40,0.2)))  
}  

### Estimating the mean   
The theoretical mean for this distrbution is 5. Figure 2 shows the distribution of the estimate of the mean from 1000 random draws of 40 exponetials. Table 1 give the summary statistics for these data. The median and mean value of these iid samples are close to the theoretical value, 5.00 and 5.02 respectively.  
  
```{r mean_estimate, echo=FALSE,fig.height=3.0, fig.width=4, fig.cap="Distribution of means of 1000\n random draws of 40 exponentials"}

mns = NULL
vrs = NULL
bins= NULL
bins2 = NULL
bins5 = NULL
set.seed(7880)
for (i in 1 : 1000) {
        tmp<-rexp(40,0.2)
        mns = c(mns, mean(tmp))
        vrs =c(vrs,var(tmp))
        bins = c(bins,ceiling((i)/100))
        bins2 = c(bins2,ceiling((i)/200))
        bins5 = c(bins5,ceiling((i)/500))
                 }
df<-data.frame(mns,bins,bins2,bins5)
hp <-ggplot(df, aes(x=mns)) + geom_histogram(binwidth=.2, colour="black", fill="#A0A0A0") + geom_vline(aes(xintercept=mean(mns)),   
               color="#FF8080", linetype="dashed", size=0.5)+ geom_vline(xintercept=5.0,color="blue", linetype="dashed", size=1)
hp <- hp + theme_light(base_size = 12, base_family = "Helvetica") + xlab("Means") +ggtitle("1000\n random draws of 40 exponentials")
hp

```
```{r mean_sum, results='asis',echo=FALSE}
#summary(mns)
print(xtable(summaryfunction(mns),caption="Summary table of means"),comment=FALSE)

```
```{r var_estimate, echo=FALSE,fig.height=3.0, fig.width=4, fig.cap="Distribution of 1000\n random draws of 40 exponentials"}
df<-data.frame(vrs)
hp <-ggplot(df, aes(x=vrs)) + geom_histogram(binwidth=5, colour="black", fill="#A0A0A0") + geom_vline(aes(xintercept=mean(vrs)),   
               color="#FF8080", linetype="dashed", size=0.5)+ geom_vline(xintercept=25.0,color="blue", linetype="dashed", size=1)
hp <- hp + theme_light(base_size = 12, base_family = "Helvetica") + xlab("Variances") +ggtitle("Distribution of Variances of 1000\n random draws of 40 exponentials")
hp
```
```{r var_sum, results='asis',echo=FALSE}
#summary(mns)
print(xtable(summaryfunction(vrs),caption="Summary table of variances"),comment=FALSE)
```

  
##Distribution  

```{r norm_estimate, echo=FALSE,fig.height=3.5, fig.width=4, fig.cap="Distribution of 1000\n random draws of 40 exponentials"}
df<-data.frame(mns)
hp <-ggplot(df, aes(x=mns-5)) + geom_histogram(binwidth=.2, colour="black", fill="#A0A0A0") + geom_vline(aes(xintercept=mean(mns-5)),   
               color="#FF8080", linetype="dashed", size=0.5)+ geom_vline(xintercept=5.0,color="blue", linetype="dashed", size=1)
hp<-
hp <- hp + theme_light(base_size = 12, base_family = "Helvetica") + xlab("Means") +ggtitle("1000\n random draws of 40 exponentials")
hp<- hp + stat_function(fun = dnorm, size = 2)
hp

```
```{r, echo=FALSE,fig.cap="Variabilty of the mean", fig.height=3}
cE<-cumsum(mns)

cP<-cE[1:1000]/seq_len(1000)
qplot(seq_len(1000),cP,xlab="Number of Averages",ylab="Estimate of the mean")
```
The **R** Code listing for this analysis can be found here:

```{r binned_mean_estimate, echo=FALSE,fig.cap=" Five distributions of 200 random draws of 40 exponentials"}
#mp <- ggplot(df, aes(x=mns)) + geom_histogram(binwidth=.2, colour="black", fill="yellow") + geom_vline(aes(xintercept=mean(mns)),   
#               color="red", linetype="dashed", size=1)+ geom_vline(xintercept=5.0,color="blue", linetype="dashed", size=1)
#gp <- mp + facet_grid(. ~ bins2)
#gp
```
